An Extension of LTL with Rules\\ and its Application to Runtime Verification

### 2019-07-15 rule-ltl.tex
Updates from Overleaf

\section{Introduction}

Runtime verification (RV)  \cite{bartocci18,havelund-rv-data-2018} very generally refers to the use of rigorous (formal) 
techniques for {\em processing} execution traces emitted by a system being observed. The purpose is typically
%, again generally viewed, 
to evaluate the state of the observed system. Processing can take numerous
forms. We focus here on {\em specification-based} runtime verification, where an execution trace is checked against a property expressed in a formal logic, in our case variants of linear temporal logic.

Linear Temporal Logic (LTL) is a common specification formalism for reactive and concurrent systems. It is often used in e.g. model checking and runtime verification. Another formalism that is used for the same purpose is finite automata, often over infinite words. This includes B\"{u}chi, Rabin, Street, Muller and Parity automata~\cite{Thomas}, all having the same expressive power. In fact, model checking of an LTL specification
is usually performed by first translating the specification into a B\"{u}chi automaton. The automata formalisms are more expressive than LTL, with a classical example by Wolper~\cite{Wolper} showing that
it is not possible to express in LTL that every even state in
the sequence satisfies $p$. This has motivated extending LTL
in various ways.
We are mainly motivated by runtime verification of executions {\em with data}, for which a first-order temporal logic is
appropriate. There, too, we show that expressiveness can be extended.

We present first an extension of  propositional LTL that is based on using additional (auxiliary) propositions, not appearing in the execution, which are dependent on the past 
of the current execution. These propositions obtain their value in a state as a function of the prefix of the execution up and including that state through a past temporal formula. This can be seen as adding
summary states to the temporal specification, resulting in a specification that is in between logic and finite automata representation. This extension, which we believe is conceptually simpler
and more intuitive than other suggested extensions, is shown to be
equivalent to the other common specification formalism:
B\"{u}chi automata, regular expressions and quantified LTL.

We then proceed to first-order LTL. We relativise Wolper's example to the first
order case to motivate the need for
extending first-order LTL. Another example is based on the inexpressiveness of the transitive closure of a relation in first order logic, where, e.g., using a relation representing neighbors in a graph is not enough for representing that there is a path between nodes.
We present an extension where auxiliary relations, not appearing in the execution, are
added. While this extension
is capable of expressing the above
properties, we show that for the first-order version, this has less 
expressive power than adding full quantification over relations. This in distinction from adding quantification to propositional LTL, where prefixing existential quantification
provides the full expressive power of, say, B\"{u}chi automata.

The extensions we proposed to the logics are in particular natural for
runtime verification of past (safety) temporal properties, since
the values of propositional and first-order LTL are based
on calculating and updating a summary of the prefix of the execution. Runtime verification is often restricted to a {\em past}  version of temporal logic~\cite{HR}, mainly because for past properties we can decide when the monitored property is violated after monitoring a finite prefix of it, while for future properties we may never know for sure. Consistent with that, 
we provide runtime verification algorithms for the past-time versions of the extended propositional and first-order
logics. We further present an extension of the 
\dejavu{} tool (publicly available on {\sf github},
hidden here because of anonymous reviewing)
\ignore{(https://github.com/havelund/dejavu)} that
realizes the extended first-order past temporal logic verification. The \dejavu{} tool~\cite{HPU,HP} allows runtime verification on past time first-order temporal logic over infinite domains (e.g., the integers, strings). It achieves efficiency by using BDD representation of enumerations of data, rather than over the data values themselves, and by using
a garbage collection algorithm.

The contributions of this paper are both theoretical and practical. On the theory side, we present and study extensions for propositional and first order linear-temporal logics and show the relations between them and existing versions. 
On the practical side, we present RV algorithms for the extended logic, and 
show how to cast them specifically in an efficient BDD-based implementation. We present some experiments performed with our implementation and report on the results.

The structure of this paper is as follows. In Section~2 we first define the syntax and semantics of propositional temporal logic. We then review some of its classical extensions and present our own extension. In Section~3 we present a runtime verification algorithm for the extended past LTL. In Section~4 we review first order linear temporal logic, present our extension and study the relationships between the presented versions of first-order LTL. In Section~5 we present a runtime verification algorithm for the past version of the extended first-order LTL. Section~6 describes the implementation and provides experimental results. Finally, Section~7 concludes the paper.

%\klaus{I think that the structure of the paper should be
%crystal clear so that the reader does not get lost. A lost reader leads to rejection. If a reader understand the structure, but perhaps not the details, the reader may still be happy.}

\vspace{1ex}
\noindent {\bf Related work.}
For propositional LTL, the logic ETL~\cite{Wolper,WVS} extends
the set of temporal operators 
by using automata (or, equivalently a right linear grammar). Each automaton defines a temporal operator, where the input letters of that automaton correspond to future temporal subformulas. Each accepted run of such an automaton defines then the places
where the subformulas need to hold. Another expressive-equivalent extension, 
in~\cite{Wolper}, adds dynamic, i.e.,
state-dependent, quantification over propositions to LTL.

Addition of relations to temporal databases is done
for aggregation~\cite{Libkin}, where relations are
calculated, based on the current state, and are used for further calculating
functions (sums, maxima) over the sequence of states. Aggregations were also used in the runtime verification tool 
{\sf MonPoly}~\cite{agrebasin}.
%
% --- other less directly related work: ---
%
Numerous other systems have been produced for monitoring
execution traces eith data against formal specifications. 
These include e.g. \cite{LOLA,larva,halle-beepbeep-ieee-12,Meredith2011,Reger2015}.

% ----------------

\iffalse
The contributions of the papers are both theoretical, and
practical. We study the extensions of propositional temporal logic (both full version and past/safety version) and suggest a restricted (and, to our opinion, an intuitive) extension that is equivalent to the power of quantified LTL or B\"{u}chi/finite automata. We then study 
The corresponding extension to first-order LTL, replacing propositions with relations, and allowing variables and constant values. We propose, for the first-order version, an extension similar to the one we propose for the propositional case. But then we show that for the first-order case, this extension has
less expressive power than adding quantification in the style of propositional temporal logic.

The extensions we suggest are in particular easy to implement, as an extension to existing propositional/first-order runtime verification. We extend our tool \dejavu, experiment with new examples that take advantage of the extension.
\fi
\vspace{1ex}
\noindent {\bf Conventions.}
\label{sec:prelim}
In this paper, we treat and present several versions of the logic LTL. To simplify
the presentation, we prefix LTL with letters as follows:

\begin{itemize}
\item{P} restricts to P{\em ast} temporal operators.
\item{F} allows F{\em irst-order} quantification over
data assigned to variables that is uniform for all states. This quantification is referred to as {\em static} quantification.
\item{Q} adds (second-order) Q{\em uantification} that is state-dependent. This is referred to as {\em dynamic} quantification.
\item{E} allows E{\em xtending} the 
set of objects that occur in the input with
new auxiliary predicates (for the propositional versions) or relations (for the first-order cases) using past temporal logic rules that define them.
\end{itemize}

%\klaus{You above introduce some concepts before they really are explained. Q and E. I understand why you do this (to have it in one place), but it may be too difficult for the reader to grasp if not "initiated".
%Especially the E.}

%\klaus{We use \ELTL{} (etc) to denote Extended LTL. 
%But \QLTL{} is also an extension of LTL. These logics 
%are all extensions of LTL. Perhaps RLTL (for Rule-based 
%LTL) would have been better ... or something. Not urgent, 
%we can always change that if we want to.}

%\klaus{The explanation of Q and E could be improved.}

%\doron{Did my best to make the letters more self explained. I think R is usually reserved in temporal logic for realtime. But if you insist, one thing that I cannot perform from here is changing the two figures where these letters occur.}

@@

The classical definition of linear temporal logic~\cite{MP} has the following syntax:
\[ \varphi ::= \true \, | \,  p \, | \, ( \varphi \wedge \varphi ) \, |  \, \neg  \varphi \, |   \, \bigcirc \varphi \, |  
\, ( \varphi  \until  \varphi ) \,|  \, \ominus \varphi |
   \, ( \varphi \ \since \ \psi ) \,  \]
where $p$ is a proposition from a finite set of propositions $P$, and $\bigcirc$, $\until$, $\ominus$, $\since$ stand for {\em next-time}, 
{\em until}, {\em previous-time} and {\em since}, respectively.

The models for LTL formulas 
are infinite sequence of states, of the form
$\sigma =  s_{1}\, s_{2}\, s_3 \ldots$,
where $s_i \subseteq P$ for each $i\geq 1$. These are the propositions that {\em hold} in that state.
%We denote by $\xi_{i}$ the suffix $s_{i} s_{i+1} s_{i+2}\ldots$ of $\sigma$. 
LTL's semantics is defined as follows:
\begin{itemize}
\item $( \sigma, {i}) \models \mathit{true}$.
\item $(\sigma, {i}) \models p$ iff $p \in s_{i}$.
\item $(\sigma, {i}) \models \neg\varphi$ iff  $(\sigma, {i}) \not\models \varphi$.
\item $(\sigma, {i}) \models ( \varphi \land \psi)$ iff $(\sigma, {i}) \models \varphi$ and $(\sigma, {i} ) \models \psi$.
\item $(\sigma, {i}) \models \bigcirc \varphi$ iff $(\sigma, {i+1}) \models \varphi$.
\item $(\sigma, {i}) \models ( \varphi  \cal{U}  \psi )$ iff for some $j$,
$j\geq i$, $(\sigma, {j}) \models \psi$,
and for each $k$, $i \leq k < j$, $(\sigma, {k}) \models \varphi$.
\item $(\sigma , i) \models \ominus \varphi$ iff $i > 1$ and $(\sigma, i-1) \models \varphi$.
\item $(\sigma , i) \models (\varphi \since \psi)$ iff there exists $j$, $1 \leq j \leq i$, such that
$(\sigma , j)  \models \psi$ and for each
$k$, $j < k \leq i$, $(\sigma , k ) \models \varphi$.
\end{itemize}
Then $\sigma \models \varphi$ when $( \sigma ,1 ) \models \varphi$.
We can use the following abbreviations:
$\false = \neg \true$, 
$(\varphi \vee \psi) = \neg ( \neg \varphi \wedge \neg \psi )$, 
$(\varphi \rightarrow \psi) = ( \neg \varphi \vee \psi)$,
$(\varphi \leftrightarrow \psi) = ((\neg \varphi \wedge \neg \psi) \vee (\varphi \wedge \psi ))$,
$\Diamond \varphi = (\true \until \varphi)$ and $\Box \varphi = \neg \Diamond \neg \varphi$,
${\bf P}\  \varphi = (\true \; \since \; \varphi)$ ({\bf P} stands for P{\em reviously}), and 
${\bf H}\  \varphi = \neg {\bf P}\ \neg \varphi$ ({\bf H} stands for 
H{\em istory}).

%%infinite sequences (finite executions can be extended by repeating the last event indefinitely to form an infinite paths).
The expressive power of different versions of propositional LTL is often compared to regular expressions
over the alphabet $\Sigma = 2^P$ and to  {\em monadic} first and second-order logics where the temporal propositions
correspond to unary predicates (over time). An overview appears in~\cite{Thomas}. Accordingly, we have the following characterizations:
LTL is equivalent to monadic first-order logic\footnote{First-order logic with unary predicates over the naturals, and the relation $<$.},  star-free regular expressions\footnote{Regular expressions but without star operator (or $\omega$).} and counter-free B\"{u}chi automata\footnote{A counter-free language is a regular language for which there is an integer $n$ such that for all words x, y, z and integers $m \geq n$ we have that $xy^m z \in L$ if and only if $xy^n z \in L$.}. In fact, restricting the temporal operators to the {\em future}  operators $\until$ and $\bigcirc$
(and the ones derived from them $\Box$ and $\Diamond$)
leaves the same expressive power and is commonly used.

%\klaus{Is that also the true for the first-order case? Doron: I have no idea. But at this point, I would not ask to ask Gabbay.}

An important subset of LTL, called here \PLTL{}, allows only past temporal
operators: $\since$, $\ominus$ and the operators derived
from them, $\mathbf{H}$ and $\mathbf{P}$. The past time logic is sometimes interpreted over finite sequences, 
where $\sigma \models \varphi$ when $( \sigma , | \sigma | ) \models \varphi$.
Alternatively,
it is also a common practice to use the past time restricted version, prefixed with  a single $\Box$ (always) operator; in this case, each of the prefixes has to satisfy $\varphi$. This later
form expresses {\em safety} LTL properties~\cite{AS} over infinite sequences. 
(When \PLTL{} is interpreted over finite sequences, its
expressive power is the same as star-free regular expressions, first-order monadic logic over finite sequences and counting-free automata.)

@@

Wolper~\cite{Wolper} showed that the expressive power of LTL is lacking with respect
to some important properties. As a canonical example, he demonstrated that it is impossible to
express that all the states with even indexes in a sequence
satisfy $p$ (this is
different than saying that $p$ alternates between $\true$ and $\false$ on consecutive states).
Wolper's example property cannot be expressed in \PLTL{} either.
In order to extend the expressiveness of LTL, Wolper suggested to use linear grammars, or, alternatively, the ability to quantify over
propositions, as shown below.

\subsubsection{Extending LTL with dynamic quantification.}
Adding quantification over propositions, as done in~\cite{Wolper}, allows writing a formula of the form
$\Exists q \, \varphi$, where
$\Exists q$ represents {\em existential} quantification over a proposition $q$ that can appear in $\varphi$.

%A model for such a formula 
%has truth assignments only over the propositions in $B$, but the propositions $A$ can appear (only) within the scope of a quantifier.
To define the semantics, let $X \subseteq P$ 
%be a subset of 
%the propositional variables appearing in the %states of the sequence $\sigma = s_1 s_2 %\ldots$
and denote $\sigma |_X = s_1 \setminus X \, s_2 \setminus X \ldots$. (Note that $\sigma |_X$ denotes projecting {\em out} the variables in $X$.)

\begin{itemize}
\item $( \sigma, i ) \models \boldsymbol{\Exists} q \, \varphi$ iff there exists
$\sigma'$ such that $\sigma' |_{\{ q \} } = \sigma$ and 
$( \sigma', i) \models \varphi$.
\end{itemize}
%Since all the auxiliary variables appear in the scope of
%quantifiers, it holds that $\sigma \models \varphi$
%(in $\varphi$, auxiliary variables appear only within
%the scope of quantifiers)
%iff $\sigma |_A \models \varphi$.
%
%\klaus{Are the last two lines above needed?}
%
Universal quantification is also allowed, where $\Forall q\, \varphi =
\neg \Exists q\, \neg \varphi$.
This kind of quantification is considered to be {\em dynamic}, since the quantified variables can have different truth values depending on the states. It is also called  {\em second-order} quantification, since the quantification establishes the {\em set} of states in which a propositional variable has the value $\true$.
Extending LTL with such quantification, the logic \QLTL{} has the same expressive power as regular expressions, full B\"{u}chi
automata, or monadic second-order logic with unary predicates over the naturals, see again~\cite{Thomas}.
In fact, it is sufficient to restrict the quantification to existential quantifiers that prefix the formula to obtain the full expressiveness of \QLTL{}~\cite{Thomas}.

Wolper's property can be expressed in \QLTL{} as follows:
\begin{equation} \label{form0} \Exists q \,
( (  \neg q \wedge \Box (q \leftrightarrow \bigcirc \neg q )  )  \wedge \Box (q \rightarrow p ) )\end{equation}

Restricting \QLTL{} to the past modalities, one obtains the logic \QPLTL{}. \QPLTL{} has the same expressive
power as regular expressions and finite automata.
Equation~(\ref{form0}) can
be rewritten in \QPLTL{} as:
\begin{equation} \label{form1} \Exists q \,
{\bf H} (( q \leftrightarrow \ominus \neg q )   \wedge (q \rightarrow p ) )\end{equation}
%In fact, it can be shown that a single existential quantifier is sufficient. 
Since $\ominus \varphi $ is interpreted as  $\false$ in the first state of any sequence, regardless of $\varphi$, then
$q$ is $\false$ in the first state. Then $q$ alternates between even and odd states.

@@

\subsubsection{Extending LTL with rules.}

We introduce now another extension of LTL, which we call \ELTL{}.
%It has a form that puts it, conceptually, in between temporal
%logic and automata: on the one hand, it is possible to simply write
%a pure LTL formula, and on the other hand, one can write a specification that resembles a finite automaton. 
As will be showed later, this extension is
is very natural for runtime verification.

We partition the propositions $P$ into
{\em auxiliary propositions} $A = \{ a_1 , \; \ldots\; , a_n \}$
and {\em basic propositions} $B$.
%A model for an \ELTL{} formula has truth assignments only over the propositions in $B$.
An \ELTL{} property $\eta$ has the following form: 
\begin{equation} \label{ELTL}
\psi \mathit{\ where\ } a_j  := \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}} \end{equation}
where each $a_j$ is a distinct auxiliary proposition from $A$,
$\psi$ is an LTL property and each $\varphi_i$ 
is a \PLTL{} property where propositions from $A$ can
only occur within the scope of a $\ominus$ operator.
The semantics can be defined as follows. We refer to $\psi$
as the {\em statement} of $\eta$ and to 
$a_j  := \varphi_j$ as a {\em rule}. In text, rules
 will be separated from each other by commas. 
\begin{tabbing}
\ \ \ $\sigma \models \eta$ \=  if there exists $\sigma'$, where
$\sigma' |_A = \sigma$ such that \\
 \> $\sigma' \models ( \psi  \wedge   \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j)) $
\end{tabbing}

\noindent
Wolper's example can be written in \ELTL{}
as follows:
\begin{equation} 
\label{form3}
\Box ( q \rightarrow p ) \, \mathrm{where} \, 
 ( q := \ominus \neg q)
\end{equation}
where $A = \{ q\}$ and $B = \{ p \}$.
The auxiliary variable $q$ is used to augment the input sequence such that each {\em odd} state will satisfy $\neg q$
and each {\em even} state will satisfy $p$.

\ELTL{} extends the set of propositional variables with new variables, whose values at a state are
functions of (i.e., uniquely defined by) the prefix of the model up to that state. The added propositions effectively summarize the past execution prefix up to and
including the current state.
This differs from 
the use of auxiliary variables in \QLTL{}, where the values assigned to the
auxiliary variables do not have to extend the 
states of the model in a unique way throughout the interpretation of 
the property over a model. The constraint that auxiliary variables appearing in the formulas $\varphi_i$ 
must occur within the scope of a $\ominus$ operator
is required to help preventing that the values
of the auxiliary variables are conflicting, as in
$a_1 := \neg a_2$ and $a_2 := a_1$, which makes a formula
that uses $a_1$ and $a_2$ uniformly $\false$. Still, one can write unsatisfiable rules, e.g., by using the rule $a_1 : = \false$.

\begin{lemma} \label{fourone}
Let $\varphi$ be an \ELTL{} formula
over the auxiliary propositions $A = \{ a_1 , \ldots , a_n \}$ and basic propositions $B$, with
a set of rules $a_j  := \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}}$. 
Let $\sigma$ be a model with states over $B$. 
Then there is at most one
model $\sigma'$ such that $\sigma' |_A = \sigma$ and $\sigma' \models \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j) $

@@

\noindent {\bf Sketch of proof.} By induction on the state index. Each auxiliary variable
is defined, via a \PLTL{} formula by the current state of the model over the values of the propositions $B$, and by the values of $A \cup B$ in previous states.  \qed
%A \PLTL{} property depends in a state only on the prefix of the model that precedes and includes that state.

\begin{theorem}
\label{theo1}
The expressive power of 
\ELTL{} is the same as \QLTL{}.
\end{theorem}
\noindent {\bf Proof.}
Each \ELTL{} formula $\eta$, as defined above,
is expressible using the following equivalent \QLTL{} formula:

\[   \Exists a_1 \ldots \Exists a_n  ( \psi  \wedge   \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j))
\]

%\klaus{How do we know that the ``other direction'' below is possible?}

\noindent
For the other direction, one can translate the \QLTL{} property first into
a second-order monadic logic formula, then to a deterministic Muller automata and then construct an \ELTL{} formula that holds for the accepting executions of this automaton. The rules of this formula encode the automata states, and the statement describes the acceptance condition of the Muller automaton.
The details of the
constructions are given in the appendix. \qed

@@

\iffalse
The use of the auxiliary variables $a_1 \ldots a_n$ and the
rules of an \ELTL{} specification 
can be viewed as representing a
state machine, in particular when the only modality that is used in the rules part is $\ominus$.
In that case, we can consider an \ELTL{} specification $\eta$ as consisting of two synchronous parts: an LTL property 
(the statement) 
and a finite automaton, whose states correspond to combinations
of truth values of the auxiliary variables, and can be referred to
within $\psi$. As a corollary of Theorem~\ref{theo1}, 
this restricted representation is also equivalent to \QLTL{}.
This representation can be useful, for RV, as explored further in this paper, because of the possibility to
incrementally calculate the values of the auxiliary variables. 
Similarly, it can be
used for extending the expressiveness of LTL model checking.

@@

We define \EPLTL{} by disallowing the future temporal operators (except an implied prefixing $\Box$) in \ELTL{}. This results in
a formalism that is equivalent to finite (B\"{u}chi) automata, where all the states except one are accepting and where the non-accepting state is a sink. 
We can use a related, but simpler construction than in Theorem~\ref{theo1} to prove the following:

@@

\item $( \epsilon ,  {\cal R} , i )
 \models p ( a ) $ if $m(p) ( a, i )$, where $a$ denotes a constant from  $d(m(p))$.

>>>>

@@

The classical definition of linear temporal logic~\cite{MP} has the following syntax:
\[ \varphi ::= \true \, | \,  p \, | \, ( \varphi \wedge \varphi ) \, |  \, \neg  \varphi \, |   \, \bigcirc \varphi \, |  
\, ( \varphi  \until  \varphi ) \,|  \, \ominus \varphi |
   \, ( \varphi \ \since \ \psi ) \,  \]
where $p$ is a proposition from a finite set of propositions $P$, and $\bigcirc$, $\until$, $\ominus$, $\since$ stand for {\em next-time}, 
{\em until}, {\em previous-time} and {\em since}, respectively.
%
The models for LTL formulas 
are infinite sequence of states, of the form
$\sigma =  s_{1}\, s_{2}\, s_3 \ldots$,
where $s_i \subseteq P$ for each $i\geq 1$. These are the propositions that {\em hold} in that state.
%We denote by $\xi_{i}$ the suffix $s_{i} s_{i+1} s_{i+2}\ldots$ of $\sigma$. 
LTL's semantics is defined as follows:
\begin{itemize}
\item $( \sigma, {i}) \models \mathit{true}$. \ \ \ \ \ 
\item $(\sigma, {i}) \models p$ iff $p \in s_{i}$. \ 
\item $(\sigma, {i}) \models \neg\varphi$ iff  $(\sigma, {i}) \not\models \varphi$.
\item $(\sigma, {i}) \models ( \varphi \land \psi)$ iff $(\sigma, {i}) \models \varphi$ and $(\sigma, {i} ) \models \psi$.
\item $(\sigma, {i}) \models \bigcirc \varphi$ iff $(\sigma, {i+1}) \models \varphi$.
\item $(\sigma, {i}) \models ( \varphi  \cal{U}  \psi )$ iff for some $j$,
$j\geq i$, $(\sigma, {j}) \models \psi$,
and for each $k$, $i \leq k < j$, $(\sigma, {k}) \models \varphi$.
\item $(\sigma , i) \models \ominus \varphi$ iff $i > 1$ and $(\sigma, i-1) \models \varphi$.
\item $(\sigma , i) \models (\varphi \since \psi)$ iff there exists $j$, $1 \leq j \leq i$, such that
$(\sigma , j)  \models \psi$ and for each
$k$, $j < k \leq i$, $(\sigma , k ) \models \varphi$.
\end{itemize}
Then $\sigma \models \varphi$ when $( \sigma ,1 ) \models \varphi$.
We can use the following abbreviations:
$\false = \neg \true$, 
$(\varphi \vee \psi) = \neg ( \neg \varphi \wedge \neg \psi )$, 
$(\varphi \rightarrow \psi) = ( \neg \varphi \vee \psi)$,
$(\varphi \leftrightarrow \psi) = ((\neg \varphi \wedge \neg \psi) \vee (\varphi \wedge \psi ))$,
$\Diamond \varphi = (\true \until \varphi)$, $\Box \varphi = \neg \Diamond \neg \varphi$,
${\bf P}\  \varphi = (\true \; \since \; \varphi)$ ({\bf P} stands for P{\em reviously}) and 
${\bf H}\  \varphi = \neg {\bf P}\ \neg \varphi$ ({\bf H} stands for 
H{\em istory}).

%%infinite sequences (finite executions can be extended by repeating the last event indefinitely to form an infinite paths).
The expressive power of different versions of propositional LTL is often compared to regular expressions
over the alphabet $\Sigma = 2^P$ and to  {\em monadic} first and second-order logic.
%, where the  propositions in the LTL formula become
%unary predicates over the naturals, and with the $<$
%relation. 
 Accordingly, we have the following characterizations:
LTL is equivalent to monadic first-order logic,  star-free regular expressions\footnote{Regular expressions without the star operator (or $\omega$).} and counter-free B\"{u}chi automata.
%\footnote{A counter-free language is a regular language for which there is an integer $n$ such that for all words x, y, z and integers $m \geq n$ we have that $xy^m z \in L$ if and only if $xy^n z \in L$.}. 
For an overview of logic and automata see~\cite{Thomas}.
Restricting the temporal operators to the {\em future}  operators $\until$ and $\bigcirc$
(and the ones derived from them $\Box$ and $\Diamond$)
maintains the same expressive power.

An important subset of LTL, called here \PLTL{}, allows only past temporal
operators: $\since$, $\ominus$ and the operators derived
from them, $\mathbf{H}$ and $\mathbf{P}$. The past-time logic is sometimes interpreted over finite sequences, 
where $\sigma \models \varphi$ when $( \sigma , | \sigma | ) \models \varphi$.
It is also a common practice to use a \PLTL{} formula, prefixed with  a single $\Box$ (always) operator; in this case, {\em each of the prefixes} has to satisfy $\varphi$. This later
form expresses {\em safety} LTL properties~\cite{AS}.
When \PLTL{} is interpreted over finite sequences, its
expressive power is the same as star-free regular expressions, first-order monadic logic over finite sequences and counting-free automata.

@@

Wolper~\cite{Wolper} demonstrated that the expressive power of LTL is lacking using
the property that all the states with even\footnote{This is
different than stating that $p$ alternates between $\true$ and $\false$ on consecutive states.} indexes in a sequence
satisfy some proposition $p$.

%Wolper's example property cannot be expressed in \PLTL{} either.
%In order to extend the expressiveness of LTL, Wolper suggested to use linear grammars, or, alternatively, the ability to quantify over
%propositions, as shown below.

\vspace{1.2ex}
\noindent {\bf Extending LTL with dynamic quantification.}
Adding quantification over propositions, suggested by Sistla in~\cite{sistla-phd-83}, allows writing a formula of the form
$\Exists q \, \varphi$, where
$\Exists q$ represents {\em existential} quantification over a proposition $q$ that can appear in $\varphi$.
%
%A model for such a formula 
%has truth assignments only over the propositions in $B$, but the propositions $A$ can appear (only) within the scope of a quantifier.
To define the semantics, let $X \subseteq P$ 
%be a subset of 
%the propositional variables appearing in the %states of the sequence $\sigma = s_1 s_2 %\ldots$
and denote $\sigma |_X = s_1 \setminus X \, s_2 \setminus X \ldots$. (Note that $\sigma |_X$ denotes projecting {\em out} the propositions in $X$.) The semantics is defined as follows:

\begin{itemize}
\item $( \sigma, i ) \models \boldsymbol{\Exists} q \, \varphi$ iff there exists
$\sigma'$ such that $\sigma' |_{\{ q \} } = \sigma$ and 
$( \sigma', i) \models \varphi$.
\end{itemize}
%Since all the auxiliary propositions appear in the scope of
%quantifiers, it holds that $\sigma \models \varphi$
%(in $\varphi$, auxiliary variables appear only within
%the scope of quantifiers)
%iff $\sigma |_A \models \varphi$.
%
%\klaus{Are the last two lines above needed?}
%
Universal quantification is also allowed, where $\Forall q\, \varphi =
\neg \Exists q\, \neg \varphi$.
This kind of quantification is considered to be {\em dynamic}, since the quantified propositions can have different truth values depending on the states. It is also called  {\em second-order} quantification, since the quantification establishes the {\em set} of states in which a proposition has the value $\true$.
Extending LTL with such quantification, the logic \QLTL{} has the same expressive power as regular expressions, full B\"{u}chi
automata, or monadic second-order logic with unary predicates over the naturals (see again~\cite{Thomas}).
In fact, it is sufficient to restrict the quantification to existential quantifiers that prefix the formula to obtain the full expressiveness of \QLTL{}~\cite{Thomas}. 
\iffalse
Wolper's property can be expressed in \QLTL{} as follows:
\begin{equation} \label{form0} \Exists q \,
( (  \neg q \wedge \Box (q \leftrightarrow \bigcirc \neg q )  )  \wedge \Box (q \rightarrow p ) )\end{equation}
\fi
Restricting \QLTL{} to the past modalities, one obtains the logic \QPLTL{}. \QPLTL{} has the same expressive
power as regular expressions and finite automata.
Wolper's property can
be rewritten in \QPLTL{} as:
\begin{equation} \label{form1} \Exists q \,
{\bf H} (( q \leftrightarrow \ominus \neg q )   \wedge (q \rightarrow p ) )\end{equation}
%In fact, it can be shown that a single existential quantifier is sufficient. 
Since $\ominus \varphi $ is interpreted as  $\false$ in the first state of any sequence, regardless of $\varphi$, then
$q$ is $\false$ in the first state. Then $q$ alternates between even and odd states.

@@

\vspace{1.5ex}
\noindent {\bf Extending LTL with rules.}
We introduce another extension of LTL, which we call \ELTL{}.
%It has a form that puts it, conceptually, in between temporal
%logic and automata: on the one hand, it is possible to simply write
%a pure LTL formula, and on the other hand, one can write a specification that resembles a finite automaton. 
As will be showed later, this extension
is very natural for runtime verification.
%
We partition the propositions $P$ into
{\em auxiliary propositions} $A = \{ a_1 , \; \ldots\; , a_n \}$
and {\em basic propositions} $B$.
%A model for an \ELTL{} formula has truth assignments only over the propositions in $B$.
An \ELTL{} property $\eta$ has the following form: 
\begin{equation} \label{ELTL}
\psi \mathit{\ where\ } a_j  := \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}} \end{equation}
where each $a_j$ is a distinct auxiliary proposition from $A$,
$\psi$ is an LTL property and each $\varphi_i$ 
is a \PLTL{} property where propositions from $A$ can
only occur within the scope of a $\ominus$ operator.
The semantics can be defined as follows. We refer to $\psi$
as the {\em statement} of $\eta$ and to 
$a_j  := \varphi_j$ as a {\em rule} (in text, rules
 will be separated  by commas).

\begin{center}
$\sigma \models \eta$ if there exists $\sigma'$, where
$\sigma' |_A = \sigma$ s.t.  
 $\sigma' \models ( \psi  \wedge   \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j)) $
 \end{center}

\noindent
Wolper's example can be written in \ELTL{}
as follows:
\begin{equation} 
\label{form3}
\Box ( q \rightarrow p ) \, \mathrm{where} \, 
  q := \ominus \neg q
\end{equation}
where $A = \{ q\}$ and $B = \{ p \}$.
The auxiliary proposition $q$ is used to augment the input sequence such that each {\em odd} state will satisfy $\neg q$
and each {\em even} state will satisfy $p$.  
%
\ELTL{} extends the set of propositions with new propositions, whose values at a state are
functions of (i.e., uniquely defined by) the prefix of the model up to that state. The added propositions summarize the past execution prefix up to and
including the current state.
This differs from 
the use of auxiliary propositions in \QLTL{}, where the values assigned to the
auxiliary propositions do not have to extend the 
states of the model in a unique way throughout the interpretation of 
the property over a model. The constraint that auxiliary propositions appearing in the formulas $\varphi_i$ 
must occur within the scope of a $\ominus$ operator
is required to prevent conflicting rules, as in
$a_1 := \neg a_2$ and $a_2 := a_1$.
%, which makes a formula
%that uses $a_1$ and $a_2$ uniformly $\false$. Still, one can write unsatisfiable rules, e.g., by using the rule $a_1 : = \false$.

\begin{lemma}[Well foundedness of auxiliary propositions] \label{fourone}
The values of the
auxiliary propositions
of an \ELTL{} formula $\eta$
are uniquely defined in a state of an execution
by the prefix of the execution upto and including that state.

@@

\noindent {\bf Proof.}
Let $\eta$ be a formula over auxiliary propositions $A$ and basic propositions $B$,
with rules
$a_j  := \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}}$.
%, with
%a set of rules $a_j  := \varphi_j \, : \, 
% {j \in \{1, \ldots , n\}}$. 
Let $\sigma$ be a model with states over $B$. 
Then there is a unique
model $\sigma'$ such that $\sigma' |_A = \sigma$ and $\sigma' \models \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j) $: inductively,
the value of each auxiliary proposition $a_j$ at the $i$th state
of $\sigma'$
is defined, via a rule $ a_j := \varphi_j$, where $\varphi_j$ is a \PLTL{} formula; 
hence it depends on
the values of the propositions $B$
in the $i$th state of $\sigma$, and on the values of $A \cup B$ in the previous states
of $\sigma'$.  \qed
%A \PLTL{} property depends in a state only on the prefix of the model that precedes and includes that state.

\begin{theorem}
\label{theo1}
The expressive power of 
\ELTL{} is the same as \QLTL{}.
\end{theorem}
\noindent {\bf Sketch of proof.}
Each \ELTL{} formula $\eta$, as defined in~(\ref{ELTL}),
is expressible using the following equivalent \QLTL{} formula:
\[   \Exists a_1 \ldots \Exists a_n  ( \psi  \wedge   \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j))
\]
%\klaus{How do we know that the ``other direction'' below is possible?}
\noindent
For the other direction, one can  first translate the \QLTL{} property  into
a second-order monadic logic formula, then to a deterministic Muller automata and then construct an \ELTL{} formula that holds for the accepting executions of this automaton. The rules of this formula encode the automata states, and the statement describes the acceptance condition of the Muller automaton.
Details are provided in an appendix. \qed

@@

\iffalse
The use of the auxiliary propositions $a_1 \ldots a_n$ and the
rules of an \ELTL{} specification 
can be viewed as representing a
state machine, in particular when the only modality that is used in the rules part is $\ominus$.
In that case, we can consider an \ELTL{} specification $\eta$ as consisting of two synchronous parts: an LTL property 
(the statement) 
and a finite automaton, whose states correspond to combinations
of truth values of the auxiliary propositions, and can be referred to
within $\psi$. As a corollary of Theorem~\ref{theo1}, 
this restricted representation is also equivalent to \QLTL{}.
This representation can be useful, for RV, as explored further in this paper, because of the possibility to
incrementally calculate the values of the auxiliary propositions. 
Similarly, it can be
used for extending the expressiveness of LTL model checking.

@@

We define \EPLTL{} by disallowing the future temporal operators (except an implied prefixing $\Box$) in \ELTL{}. This results in
a formalism that is equivalent to a B\"{u}chi automata, where all the states except one are accepting and where the non-accepting state is a sink. 
We can use a related, but simpler construction than in Theorem~\ref{theo1} to prove the following:

@@

\item $( \epsilon ,  {\cal R} , i )
 \models p ( a ) $ if $m(p) ( a, i )$, where $a$ denotes a constant from  $dom (m(p))$.
